The First Draft

### 2023-04-25 ArticleDraft.Rmd
Final first draft

---
title: "The First Draft"
authors:
  - name: "Mehrdad Samadishadlou"
    address: "Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran"
date: "`r Sys.Date()`"
link-citation: yes
bibliography: For Article Refrence.bib
output:
  bookdown::pdf_book:
  bookdown::word_document2: default
#    base_format: rticles::bioinformatics_article
---

```{r setup, include=FALSE, cache.rebuild=TRUE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, captioner, bundesligR, stringr)
```

```{r , echo=F}
# What to do next:

# Generating information related to the machine learning first layer using welcome file.

```

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO) database (<https://www.ncbi.nlm.nih.gov/geo/>). In order to come by sufficient classification power between MI samples and others, the sample size should be relatively large. Therefore, GSE59867 for MI and CAD samples, and GSE56609 and GSE54475 for healthy samples were selected. All samples were produced using Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. Only healthy, stable CAD and early stage MI samples were selected from these datasets for further analysis. The basic information for the three datasets evaluated in the current study is provided in Table 1.

|          | Platform | Healthy Control | CAD Control | MI  | Reference |
|----------|----------|-----------------|-------------|-----|-----------|
| GSE59867 | GPL6244  | \-              | 46          | 111 | [@67]     |
| GSE56609 | GPL6244  | 46              | \-          | \-  | [@09]     |
| GSE54475 | GPL6244  | 5               | \-          | \-  | [@75]     |

: Basic information of the GEO microarray datasets.

@@

Raw data (CEL files) of all datasets were downloaded from the GEO and pre-processed using the fRMA package [@fRMA]. fRMA allows to preprocess individual microarray samples and combine them consistently for analysis. For each dataset, background correction was performed using the RMA algorithm and then it was quantile normalized based on the reference distribution. During summarization, batch effects were removed and variances of the gene expressions were estimated by taking into account these probe-specific effects. For those multiple probe sets matched to the identical gene, the mean log fold change was retained. This way, fRMA can be seen as a batch effect removal technique for different datasets that produced identical microarray platforms. Thus, to become sure about batch effect removal, the principal component analysis and the relative log expression of train samples were plotted before and after fRMA [@BER].

@@

The barcode algorithm proposed by McCall et al. [@Barcode] transforms the actual expression values into binary barcode values. Huge sets of samples were collected and normalized using fRMA for several platforms. The distribution of the expressed and unexpressed observed intensities for each gene is estimated using these normalized sets. Genes are deemed expressed (and their value coded to 1) or unexpressed (and their value coded to 0) according to the following equation:

@@

where $x_{ij}$ is the normalized intensity of gene $i$ in sample $j$, $C$ is a user-defined parameter, $\sigma^{ne}$ is the standard deviation of the non-expressed distribution and $\mu^{ne}$ is the mean of the non-expressed distribution. The barcode representation of a sample is a vector of ones and zeros denoting which genes are estimated to be expressed (ones) and unexpressed (zeros). The barcode algorithm was implemented by the barcode function in the R fRMA package, and the default value of $C$ was used.

To determine if the expressed ratios differed in the MI group versus the healthy control group, Fisher's exact test for individual genes was carried out upon the barcode values. Genes with a false discovery rate (FDR) of $< 0.05$, which was calculated through the Benjamini-Hochberg (BH) procedure to adjust for multiple testing issues, were considered as differentially expressed genes.

The same procedures were conducted on CAD versus healthy controls as well as MI versus CAD group to find the DEGs in between them.

@@

Using the R clusterProfilter package, the Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analysis and Gene Ontology (GO) functional annotation were carried out on the differentially expressed genes. The GO analysis included biological process (BP), cellular component (CC) and molecular function (MF) categories. An adjusted P \< 0.05 was considered to indicate a statistically significant difference. Enrichments were conducted on the MI-healthy and CAD-healthy DEGs. In these analyses, all default parameters were used.

@@

The machine learning analysis was performed using Python software, ver. 3.9, numpy [@numpy], pandas [@pandas] and Scikit-Learn packages [@scikit-learn]. Whenever hyper-tuning was needed the scikit-opt package [@scikitopt] was used. In all ML analysis, the datasets were divided into train and test sets by 0.7:0.3 ratio and all reported results are the average of a 10-fold cross-validation.

Two different approaches were used for selecting microRNAs to train models with. The first approach was using the microRNAs that are differentially expressed. In the second approach microRNAs were selected by their individual AUC-ROC. Having the result of these two different approaches can provide an informative comparison between the predictive capabilities of a set of microRNAs selected with different logic.

### microRNAs in DEGs

In this approach a two layer architecture was deployed to the data in order to maximize the prediction values. The first layer predicted whether a sample is healthy or not, and the second layer separated MI from CAD in the samples which were predicted as not healthy in the first layer. To this end, a distinct ML model was trained for each layer. Since there are limited numbers of microRNAs in DEGs both layers were trained with all of them. For further comparison with the models performance ROC curve of each microRNA for classifying healthy and not-healthy, as well as CAD and MI was generated using a simple Logistic Regression model.

#### First layer for seperating healthy and not-healthy samples:

An SVM model using RBF kernels was hyper-tunned and trained using all microRNAs in DEGs. In order to handle the imbalance in the number of samples  (51 for healthy and 157 for not-healthy group), sample weight for not-healthy samples was set to 0.5.

For the sake of reaching the highest classification performance using the set of microRNAs, different models were investigated. To do so, SVM (with linear, polynomial, and RBF kernels), Logistic Regression (LR), Random Forests (RF), k-Nearest Neighbor (kNN), Gradient Boosting (GB), XGBoost (XGB) and Decision Tree (DT) models were trained. All models were trained with their pre-set parameters with 10-fold cross-validation.

Criteria for choosing the best model was the highest accuracy and AUC on the test set. The best model was hyper-tuned with the scikit-opt package [@scikitopt] in order to get the best predictive performance.

### microRNAs with the highest AUC-ROC

Like the previous approach, a two layers strategy was conducted. The first layer for classifying samples to healthy and not-healthy and the second layer for separating MI and CAD samples. However, to keep the number of microRNAs as low as possible microRNAs will be selected from the second layer (which are the microRNAs with the best performance in MI/CAD separation) and then their performance will be evaluated in the first layer.

AUC-ROC of all microRNAs for classifying MI and CAD samples has been calculated. For finding the number of microRNAs with the highest predictive values, the microRNAs with the highest individual AUC-ROC are added to the set one-by-one and the AUC-ROC for the set is calculated. The set with the highest AUC-ROC has been selected as the set for the following steps. The ROC curves for each selected microRNA for separating healthy samples from not-healthy and MIs from CADs were also plotted for further comparison.

#### First layer for finding healthy and not-healthy samples:

An SVM model with RBF kernel was trained with the selected set of microRNAs. Moreover, the model was hyper-tuned for finding the hyper-parameters with the best AUC and accuracy. The ROC curve and confusion matrix for the best model was reported.

The selected number of microRNAs was used to train different algorithms for finding the best model. As the previous approach, SVM (with linear, polynomial, and RBF kernels), LR, RF, kNN, GB, XGB and DT were trained. All models were trained with their pre-set parameters with 10-fold cross-validation. The models with the highest AUC-ROC and accuracy on the test set were selected and hyper-tuned with the scikit-opt package [@scikitopt]. The ROC curve and confusion matrix for the best model was reported.

# Results and Discussion

The PCA plot of the train samples were shown in Fig\@ref(fig:PCA). As it is clear, there was a complete separation between healthy samples and CAD and MI samples in primary data which also remains after conducting fRMA on the data.

The RLE plot presented in Fig\@ref(fig:PCA) validates batch effect removal. For an efficient batch effect removal method, the individual boxplots will be all distributed around 0 in the RLE plot, and inter-quantile distances would be greater than 0.1 [@BER]. The mentioned criteria was not met in primary data, but was met after conducting fRMA algorithm.

```{r PCA, echo=F, out.width="100%", fig.cap="PCA and RLE plot for all samples before and after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

```{r , echo=F}
M_H <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs MI-Healthy.txt", check.names = F)

C_H <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs CAD-Healthy.txt", check.names = F)
```

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between the MI and the healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in MI compared to the healthy controls. For CAD and healthy groups there were `r dim(C_H)[1]` DEGs, `r sum(C_H$logFC > 0)` of them were up-regulated and `r sum(C_H$logFC < 0)` of them were down-regulated in CAD samples in comparison with healthy samples. For MI and CAD group the number of DEGs are `r dim(M_C)[1]`, and the number of up- and down-regulated genes in MI samples in comparison with CAD samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)`, respectively. All these data is summarized in table 2.

|                 | No. of DEGs     | no. of up-regulated DEGs | no. of down-regulated DEGs | microRNAs                           |
|---------------|---------------|---------------|---------------|---------------|
| MI vs. Healthy  | `r dim(M_H)[1]` | `r sum(M_H$logFC > 0)`   | `r sum(M_H$logFC < 0)`     | hsa-miR-186, hsa-miR-21, hsa-miR-32 |
| CAD vs. Healthy | `r dim(C_H)[1]` | `r sum(C_H$logFC > 0)`   | `r sum(C_H$logFC < 0)`     | hsa-miR-186, hsa-miR-21, hsa-miR-32 |
| MI vs. CAD      | `r dim(M_C)[1]` | `r sum(M_C$logFC > 0)`   | `r sum(M_C$logFC < 0)`     | hsa-miR-186                         |

: Total DEGs and differentially expressed microRNAs.

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on MI-healthy and CAD-healthy DEGs.

For MI versus healthy samples, GO enrichment analysis in the BP category, suggested that the DEGs were enriched in "immune response-regulating signaling pathway", "lymphocyte differentiation", "immune response-regulating cell surface receptor signaling pathway", and "leukocyte activation involved in immune response" (Fig\@ref(fig:MIHEnrich)A). In the CC category the DEGs were enriched in "secretory granule membrane", "azurophil granule", "ficolin-1-rich granule", "tertiary granule", and "ficolin-1-rich granule membrane" (Fig\@ref(fig:MIHEnrich)B).  In the MF category, the DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Fig\@ref(fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Fig\@ref(fig:MIHEnrich)D).

```{r MIHEnrich, echo=F, out.width="95%", fig.cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the MI and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich MI-Healthy.pdf")
```

The enrichment results for DEGs of CAD versus healthy samples were as follows. In the BP category, GO enrichment suggested that the DEGs were enriched in "positive regulation of defense response", "positive regulation of innate immune response", "mononuclear cell differentiation", and "positive regulation of response to external stimulus" (Fig\@ref(fig:CADHEnrich)A). In the CC category the DEGs were enriched in "azurophil granule", "ficolin-1-rich granule", and "ficolin-1-rich granule membrane" (Fig\@ref(fig:CADHEnrich)B).  In the MF category, the DEGs were involved in "lipoprotein particle receptor binding" and "NF-$\kappa$B binding" (Fig\@ref(fig:CADHEnrich)C). KEGG pathway analysis indicated that the  DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Fig\@ref(fig:CADHEnrich)D).

```{r CADHEnrich, echo=F, out.width="95%", fig.cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the CAD and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich CAD-Healthy.pdf")
```

### microRNAs in DEGs

>>>>

---
title: "The First Draft"
authors:
  - name: "Mehrdad Samadishadlou"
    address: "Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran"
date: "`r Sys.Date()`"
link-citation: yes
bibliography: [For Article Refrence.bib, discussion.bib, Introduction.bib]
output:
  bookdown::word_document2: default
  bookdown::pdf_book:
#    base_format: rticles::bioinformatics_article
always_allow_html: true
---

```{r setup, include=FALSE, cache.rebuild=TRUE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, captioner, bundesligR, stringr, kableExtra)
```

# Introduction

At present, cardiovascular diseases (CVDs) are the leading cause of human mortality with 32% of all global deaths. It is estimated that about 85% of CVDs mortality were diagnosed with myocardial infarction (MI) [@CVD]. MI is an acute coronary syndrome with sudden blockage and stenosis of the coronary artery, and subsequent myocardial ischemia, leading to extensive cardiomyocyte damage and necrosis [@Macro].

Over the last 50 years, numerous attempts have been collected to use biomarkers to facilitate diagnosis, assess risk, follow-up therapy, and determine therapeutic efficacy in CVDs candidates. Based on the released guidelines, cardiac troponins (cTns) are used as a highly-sensitive and accurate approach for the detection of myocardial ischemia. Despite the inherent advantages, the high-rate sensitivity of cTn-based assays has also led to more false positive results [@17], which do necessitate the advent and development of new modalities with pathological values. To improve diagnostic value upon existing CVD biomarkers, the combination of complementary biological markers, such as microRNAs (miRNAs) and other genetic factors, is proposed. Previous data support the notion that miRNAs exhibit the great potential to be used as alternative biomarkers in CVD detection and follow-up [@miR]. It is suggested that miRNAs possess 18-22 nucleotides and can play a crucial role in the regulation of gene expression. Evidence point to the fact that miRNAs are involved in the pathogenesis of cardiac tissue injury and be as theranostics in terms of CVDs. These elements can easily circulate in biofluids with both diagnostic as well as prognostic values [@m1]. Several biological activities such as angiogenesis, cardiomyocyte growth and contractility, lipid metabolism, plaque formation, and cardiac rhythm are regulated by miRNAs [@m2]. It is postulated that the function and diagnostic properties of miRNAs are beyond the myocardium in CVD patients. To be specific, the expression of miRNAs can vary in different biofluids and cell components such as serum and peripheral blood mononuclear cells (PBMCs) [@36].

PBMCs are a fraction of white blood cells (WBCs), including monocytes, lymphocytes, macrophages, and other cells belonging to the immune system [@PBMC-miR]. Emerging data have indicated that PBMCs can be used as a valid source of biomarkers for monitoring various pathological conditions. Of note, the alteration of mRNAs and miRNAs under pathological conditions gives us valuable information about different kinds of disorders. PBMCs could recapitulate the conditions of the target tissues, thus, providing a highly sensitive and specific source of biomarkers [@Meder6]. Commensurate with these conditions, these cells are repositories of dysregulated genes and miRNAs expression profiles in CVDs related to control conditions [@PBMC-miR; @Meder6].

In recent years, the advent and use of machine learning (ML) is an exciting prospect for advancing scientific discoveries. Although the concept of ML and its initial algorithms were conceived many years ago, recent improvements in computing power and access to vast amounts of data have shown that ML techniques outperform classical statistical methods in various fields. Furthermore, the progress made in omics technologies has enabled the analysis of massive and intricate biological data sets, consisting of hundreds to thousands of samples, which makes it possible for ML to extract valuable biological information from such data [@ML-Bio]. Therefore, ML offers novel techniques to integrate and analyze the various omics data enabling the discovery of de novo biomarkers. These biomarkers help us in accurate disease prediction, patient stratification, and finding new therapeutics [@ML].

In this study, we aimed to identify potential miRNA biomarkers for MI patients by combining and analyzing three different microarray datasets from PBMCs. It is suggested that the integration of omics data with bioinformatics and ML techniques could be a promising tool in the discovery of new and more accurate biomarkers for monitoring CVDs. Besides, this approach can deepen our vision into the underlying mechanisms of CVDs and aid in the development of valid theranostic tools, and patient stratification.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO) database (<https://www.ncbi.nlm.nih.gov/geo/>). To obtain sufficient classification power between MI, healthy and CAD samples, a relatively large sample size was required. Therefore, GSE59867 for MI and CAD samples, and GSE56609 and GSE54475 for healthy samples were selected. All samples were produced using Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. Only healthy, stable CAD and early-stage MI samples were selected from these datasets for further analyses. The basic information for the three datasets evaluated in the current study is provided in Table \@ref(tab:datasets).

```{r, datasets, echo=FALSE}
df <- data.frame(Dataset = c("GSE59867", "GSE56609", "GSE54475"),
                 Platform = c("GPL6244", "GPL6244", "GPL6244"),
                 Healthy = c("-", 46, 5),
                 CAD = c(46, "-", "-"),
                 MI = c(111, "-", "-"),
                 Refrence = c("[@67]", "[@09]", "[@75]"))

kbl(df, format = "markdown", booktabs = T, align = "c", linesep = c(""), caption = "Basic information of the GEO microarray datasets.")
```

@@

Raw data (CEL files) of all datasets were downloaded from the GEO and pre-processed using the fRMA package [@fRMA]. fRMA allowed to pre-process of individual microarray samples and combining them consistently for analysis. For each dataset, background correction was performed using the RMA algorithm and then it was quantile normalized based on the reference distribution. During summarization, batch effects were removed and variances of the gene expressions were estimated by taking into account these probe-specific effects. For those multiple probe sets matched to the identical gene, the mean log fold change was retained. Therefore, fRMA can be seen as a batch effect removal technique for different datasets that are produced by identical microarray platforms. Thus, to ensure batch effect removal, the principal component analysis and the relative log expression of train samples were plotted before and after fRMA [@BER].

@@

The barcode algorithm proposed by McCall et al. [@Barcode] transformed the actual expression values into binary barcode values. Huge sets of samples were collected and normalized using fRMA for several platforms as well as for Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. The distribution of the expressed and unexpressed observed intensities for each gene is estimated using these normalized sets. Genes were considered expressed (and their value coded to 1) or unexpressed (and their value coded to 0) according to the following equation:

@@

where $x_{ij}$ is the normalized intensity of gene $i$ in sample $j$, $C$ is a user-defined parameter, $\sigma^{ne}$ is the standard deviation of the non-expressed distribution, and $\mu^{ne}$ is the mean of the non-expressed distribution. The barcode representation of a sample is a vector of ones and zeros denoting which genes are estimated to be expressed (ones) and unexpressed (zeros). The barcode algorithm was implemented by the barcode function in the R fRMA package, and the default value of $C$ was used.

To determine if the expressed ratios differed in the MI group versus the healthy control group, Fisher's exact test for individual genes was carried out upon the barcode values. Genes with a false discovery rate (FDR) of $< 0.05$, which was calculated through the Benjamini-Hochberg (BH) procedure to adjust for multiple testing issues, were considered as differentially expressed genes. The same procedures were conducted on CAD versus healthy controls as well as MI versus CAD group to find the DEGs between them.

@@

Using the R clusterProfiler package, the Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analysis and Gene Ontology (GO) functional annotation were carried out on the differentially expressed genes. The GO analysis included biological process (BP), cellular component (CC) and molecular function (MF) categories. An adjusted p-value of less than 0.05 was considered to indicate a statistically significant difference. Enrichments were conducted on the MI-healthy and CAD-healthy DEGs. In these analyses, all default parameters were used.

## ML procedure

The ML analysis was performed using Python software, ver. 3.9, Numpy [@numpy], pandas [@pandas], and Scikit-Learn packages [@scikit-learn]. Whenever hyper-tuning was needed, the scikit-opt package [@scikitopt] was used. In all ML analyses, the datasets were divided into train and test sets by a 0.7:0.3 ratio and all reported results are the average of 10-fold cross-validation.

Two different approaches were used for selecting miRNAs for model training. The first approach was using the miRNAs that are differentially expressed. In the second approach, miRNAs were selected by their individual AUC-ROC. Having the result of these two different approaches can provide an informative comparison between the predictive capabilities of sets of miRNAs selected with different logics.

### miRNAs in DEGs

In this approach, a two-layer architecture was deployed to the data to maximize the prediction values. The first layer predicted whether a sample is healthy or not, and the second layer separated MI from CAD in the samples which were predicted as not healthy in the first layer. To this end, a distinct ML model was trained for each layer. Since there is a limited number of miRNAs in DEGs, both layers were trained with all of them. For further comparison with the models' performance, the ROC curve of each miRNA for classifying healthy and not-healthy, as well as CAD and MI, were generated using a Logistic Regression model.

#### First layer for isolation of healthy and not-healthy samples:

An SVM model using RBF kernels was trained and  hyper-tuned using all miRNAs in DEGs. To handle the severe imbalance in the number of samples (51 for the healthy and 157 for the not-healthy groups), the sample weight for the healthy and the not-healthy samples were set to 1 and 0.5, respectively.

For the sake of reaching the highest classification performance using the set of miRNAs, different models were investigated. To do so, SVM (with linear, polynomial, and RBF kernels), Logistic Regression (LR), Random Forests (RF), k-Nearest Neighbor (kNN), Gradient Boosting (GB), XGBoost (XGB) and Decision Tree (DT) models were trained. All models were trained with their pre-set parameters with 10-fold cross-validation. The criteria for choosing the best model were the highest accuracy and AUC on the test set. The best model was hyper-tuned with the scikit-opt package [@scikitopt] to get the best predictive performance.

### miRNAs with the highest AUC-ROC

Like the previous approach, a two-layers strategy was conducted. The first layer classified samples into healthy and not-healthy, and the separated MI and CAD samples. However, to keep the number of miRNAs as low as possible miRNAs were selected from the second layer (which are the miRNAs with the best performance in MI/CAD separation), and then their performance wase evaluated in the first layer. AUC-ROC of all miRNAs for classifying MI and CAD samples were calculated. To find the number of miRNAs with the highest predictive values, the miRNAs with the highest individual AUC-ROC were added to the set one by one, and the AUC-ROC for the set was calculated. The set with the highest AUC-ROC was selected for the following steps. The ROC curves for each selected miRNA for separating healthy samples from not-healthy and MIs from CADs were also plotted for further comparison.

#### First layer for the detection of healthy and not-healthy samples:

An SVM model with an RBF kernel was trained using the selected set of miRNAs. Additionally, the model was hyper-tuned to find the hyper-parameters for the highest AUC and accuracy. The ROC curve and confusion matrix for the best model were reported.

The selected miRNAs set was used to train different algorithms to find the best model. Similar to the previous approach, SVM (with linear, polynomial, and RBF kernels), LR, RF, kNN, GB, XGB, and DT were trained. All models were trained with their pre-set parameters using 10-fold cross-validation. The models with the highest AUC-ROC and accuracy on the test set were selected and hyper-tuned using the scikit-opt package [@scikitopt]. The ROC curve and confusion matrix for the best model were reported.

# Results

The PCA plots of the training samples are shown in Figure \@ref(Fig:PCA)A and B. As shown, healthy samples were separated from CAD or MI samples in primary data and also after conducting fRMA. In the RLE plot, there was a distinct difference between dataset means for all samples before conducting fRMA  (Figure \@ref(Fig:PCA)C). All datasets were rearranged around 0 in the RLE plot after conducting fRMA (Figure \@ref(Fig:PCA)D). Moreover, there was a clear change in inter-quantile distances, but the values still were over 0.1.

```{r PCA, echo=F, out.width="100%", Figure .cap="PCA and RLE plot for all samples before and after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

```{r , DEGstab, echo=F}
M_H <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs MI-Healthy.txt", check.names = F)

C_H <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs CAD-Healthy.txt", check.names = F)

df <- t(data.frame(`MI vs. Healthy` = c(dim(M_H)[1], sum(M_H$logFC > 0), sum(M_H$logFC < 0), "hsa-miR-186, miR-21, miR-32"),
       `CAD vs. Healthy` =  c(dim(C_H)[1], sum(C_H$logFC > 0), sum(C_H$logFC < 0), "hsa-miR-186, miR-21, miR-32"), 
       `MI vs. CAD` = c(dim(M_C)[1], sum(M_C$logFC > 0),  sum(M_C$logFC < 0), "hsa-miR-186")))

colnames(df) <- c("DEGs",  "up-regulated DEGs", "down-regulated DEGs", "miRNAs")
rownames(df) <- c("MI vs. Healthy", "CAD vs. Healthy", "MI vs. CAD")

kbl(df, booktabs = T, align = "c", linesep = c(""), caption = "Total, up-, and down-regulated DEGs and differentially expressed miRNAs.")%>% 
  kable_styling(full_width = T) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(5, width = "4cm")
```

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between the MI and the healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in MI compared to the healthy controls. In CAD and healthy groups comparison, we found `r dim(C_H)[1]` DEGs, of which `r sum(C_H$logFC > 0)` and and `r sum(C_H$logFC < 0)` DEGs were up- and `r down-regulated, respectively in CAD samples. In the MI and CAD groups, the number of DEGs was `r dim(M_C)[1]`, and the number of up- and down-regulated genes in MI samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)`, respectively in comparison with CAD samples. These data are summarized in Table \@ref(tab:DEGstab).

```{r venn, echo=F, out.width="60%", Figure .align='center', Figure .cap="Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD comparison."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/venny.png")
```

The Venn diagram  in Figure \@ref(Fig:venn) shows that CAD and MI samples share a majority of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs of CAD/healthy, 531 genes were common which is 62% of MI/healthy DEGs and 79% of CAD/healthy DEGs.

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on MI-healthy and CAD-healthy DEGs. In MI versus healthy samples, GO enrichment analysis in the BP category, suggested that the DEGs were enriched in "immune response-regulating signaling pathway", "lymphocyte differentiation", "immune response-regulating cell surface receptor signaling pathway", and "leukocyte activation involved in immune response" (Figure \@ref(Fig:MIHEnrich)A). In the CC category, the DEGs were enriched in "secretory granule membrane", "azurophil granule", "ficolin-1-rich granule", "tertiary granule", and "ficolin-1-rich granule membrane" (Figure \@ref(Fig:MIHEnrich)B). In the MF category, the DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Figure \@ref(Fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Figure \@ref(Fig:MIHEnrich)D).

```{r MIHEnrich, echo=F, out.width="95%", Figure .cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the MI and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich MI-Healthy.pdf")
```

The enrichment results for DEGs of CAD versus healthy samples were as follows. In the BP category, GO enrichment suggested that the DEGs were enriched in "positive regulation of defense response", "positive regulation of innate immune response", "mononuclear cell differentiation", and "positive regulation of response to external stimulus" (Figure \@ref(Fig:CADHEnrich)A). In the CC category, the DEGs were enriched in "azurophil granule", "ficolin-1-rich granule", and "ficolin-1-rich granule membrane" (Figure \@ref(Fig:CADHEnrich)B). In the MF category, the DEGs were involved in "lipoprotein particle receptor binding" and "NF-$\kappa$B binding" (Figure \@ref(Fig:CADHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Figure \@ref(Fig:CADHEnrich)D).

```{r CADHEnrich, echo=F, out.width="95%", Figure .cap="Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the CAD and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/enrich CAD-Healthy.pdf")
```

@@

### miRNAs in DEGs

Among all DEGs, just miR-186, miR-32, and miR-21 were detected as differentially expressed miRNAs. The expression profile of these three miRNAs is presented in Figure \@ref(Fig:DEMexp). Additionally, The ROC curves of each miRNA for each layer are presented in Figure \@ref(Fig:miRROC). Using the logistic regression model the AUC for miR-21, miR-32, and miR-186 was 0.99, 1, and 0.91 respectively  (Figure \@ref(Fig:miRROC)A). Besides, the accuracy of each miRNA for classifying the samples into healthy and not-healthy groups on the test set was 0.92, 0.98, and 0.89 for miR-21, miR-32, and miR-186, respectively. Moreover, the ROC curve of each miRNA for classifying MI and CAD samples was presented in Figure \@ref(Fig:miRROC)B. For miR-21, miR-32, and miR-186, the AUC and accuracy on the test set were 0.85; 0.7; and 0.82, and  0.78; 0.67; and 0.74, respectively.

```{r, mirExptable, echo=FALSE}

# data
df <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/microRNAs logFC.txt", sep = "\t", check.names = F, header = T)

kbl(df, booktabs = T, align = "c",linesep = c(""), digits = c(1, 60), caption = "Target miRNAs log fold-change and adjusted p-values for CAD samples relative to healthy, MI samples relative to healthy, and MI samples relative to CAD.") %>%
  kable_styling(full_width = T) %>%
  add_header_above(c(" " = 1, "CAD/Healthy" = 2, "MI/Healthy" = 2, "MI/CAD" = 2))
```

```{r DEMexp, echo=F, out.width="50%", Figure .align='center', Figure .cap="Expression profile of miR-186, miR-21, and miR32 in Healthy, CAD, and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEGs healthy not-healthy/DEMs Expression.pdf")
```

```{r miRROC, echo=F, out.width="95%", Figure .align='center', Figure .cap="ROC curve for single miRNAs on test set classification for (A) healthy and not-healthy samples and (B) CAD and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/miRs ROCs.pdf")
```

#### First layer for healthy not-healthy isolation:

Although single miRNAs had acceptable performance, their predictive value could be improved even further by using them as a set. The ROC curve for the SVM model with an RBF kernel trained with all three miRNAs is presented in Figure \@ref(Fig:DEMROC)A. The model had a better performance in classification than single miRNAs.The AUC for the model is 1, and its accuracy on the test set was also 1. In Figure \@ref(Fig:DEMCF)A, the confusion matrix for the model is presented.

```{r DEMROC, echo=F, out.width="95%", Figure .align='center', Figure .cap="ROC curve for miRNAs in DEGs on test set classification for (A) healthy and not-healthy samples and (B) CAD and MI samples."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEMs ROCs h not h cad mi.pdf")
```

```{r DEMCF, echo=F, out.width="90%", Figure .align='center', Figure .cap="Confusion matrix on the test set for (A) An SVM model with RBF kernel for healthy and not-healthy and (B) An SVM model with linear kernel for CAD and MI samples classification."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/DEMs Conf. matrix.pdf")
```
